\Large Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the\\Parametric Bootstrap

### 2017-03-25 TeX/paper.tex
Coverage_probability_plots

\section*{Efficiency of MPLE/MCMCMLE}
Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al. \cite{vanDuijn.2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE. Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\[0.3cm]
\noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$. For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution. We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMCMLE of these simulated networks. For every single simulated network the MCMCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 
$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$
implying that the smaller the RMSE, the more accurate is the estimator. Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMCMLE to the MPLE. As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance. Figure \ref{rmse} visualizes the results of the simulation study. The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.

@@

\section*{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMCMLE if the sample size is not chosen large enough. The main reason why the MCMCLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}. By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.
We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. \\[0.3cm]
For the simulation study, we determine the MPLE for the model and treat these estimates as the networks 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks we calculate the confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the MPLE estimates to obtain 95\% bootstrap confidence intervals. Similar as for the MCMCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval. Figures ref1 and ref2 visualize the coverage probabilities for each of the three methods for both networks.

\begin{figure}[h]
\begin{center}
\includegraphics[scale=.4]{Coverage_faux_mesa_high}
\caption{The Coverage Probability results of the Faux Mesa High network for bootstrapped MPLE, MCMCMLE and logistic regression }
\label{rmse}
\end{center}
\end{figure}

\item include magnolia plot
\item Describe plots
\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
\item can easily be parallelized/ using multiple cores -> increases speed
\item this method is also able to detect degeneracy

\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit, especially if the size of the examined network is extremely large.

@@

@@

 \section*{To determine}

 \begin{enumerate}
 \item where to include Boltzmann machine application
\end{enumerate}

>>>>

\section*{Efficiency of MPLE/MCMCMLE}
Even though the MCMCMLE is in general favored over the MPLE method there are also cases where the MPLE comes in handy. Foremost, MPLE is quick and simple, since estimation can be done by basic logistic regression and does not exert elaborate MCMC methods. Furthermore, as mentioned in the previous chapter the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Strauss et al /cite, Hyvarinen /cite, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE becomes an option to computationally intensive MCMC methods.\\
However, another major advantage of the MCMCMLE method is, as shown by van Duijn et al. \cite{vanDuijnetal2009}, that the MCMCMLE is a more efficient estimator than the MPLE, meaning that the variance of the MCMCMLE is in general lower than the variance of the MPLE. Nevertheless, as shown by Desmarais and Cranmer \cite{Desmarais.2012} the MPLE outperforms the MCMCMLE if the number of simulated networks used to approximate the likelihood is not chosen large enough. It is even more remarkable that the number of simulated networks needed for the MCMCMLE, in order to surpass the MPLE increases as the size of the network increases. This means that for very large networks it becomes difficult to determine the number of simulated networks needed in order for the MCMCMLE to outperform the MPLE. In other words, the larger the network of interest is, the larger one has to choose the number of simulated networks in order to justify the time intensive MCMCMLE approach.\\[0.3cm]
\noindent In order to demonstrate this disadvantage of the MCMCMLE we conduct a simulation study using Goodreau's Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al \cite{Resnicketal1997}.\\[0.3cm]
For both networks we first calculate the MCMCMLE and treat the estimated coefficients as the network's true values $\theta$. For both networks we take the same parametrization, using the number of edges, the nodal attribute for gender and the geometrically weighted edgewise shared partners (gwesp)(\cite{Hunter.2006}) distribution. We simulate 500 new networks using the 'true' coefficients and estimate the MPLE as well as the MCMCMLE of these simulated networks. For every single simulated network the MCMCMLE calculation is being repeated several times using 1500, 800, 400, 200, 100, 50 and 25 simulated networks in order to approximate the likelihood.
Based on these results, we compute the root mean square error, which is a measure of the accuracy of an estimator, combining both, the bias and the variance. Mathematically written, the RMSE for an estimator $\hat{\theta}$ is defined as 
$$RMSE = \sqrt{\sum_{i=1}^{500}(\theta - \hat{\theta}_i)^2}$$
implying that the smaller the RMSE, the more accurate is the estimator. Since the MCMCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMCMLE to the MPLE. As a result, a negative value indicates a better MCMCMLE performance, while a positive value indicates a better MPLE performance. Figure \ref{rmse} visualizes the results of the simulation study. The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network.

@@

\section*{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMCMLE if the sample size is not chosen large enough. The main reason why the MCMCLE is still widely preferred is that in contrast to the MPLE, it does not underestimate the variance of its estimates \cite{vanDuijnetal2009}. By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (Y)$ is the sufficient statistic. This allows us to work in the framework of exponential family distributions and as a consequence, leads to the conclusion that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLEs and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that only comprise the true value in less than 75\% instead of the demanded 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\begin{figure}[h]
\begin{center}
\includegraphics[scale=.45]{Coverage_Mesa_Magnolia}
\caption{The Coverage Probability results of the Faux Mesa High network (left) and of the Faux Magnolia High network (right) for bootstrapped MPLE, MCMCMLE and logistic regression }
\label{coverage}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[scale=.40]{Bias_plots}
\caption{The boxplots visualize the bias $(\hat{\theta}-\theta)$ over the 500 iterations for the Faux Mesa High network (left) and the Faux Magnolia High network (right).}
\label{bias}
\end{center}
\end{figure}
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we are going to apply a technique, referred to as \textit{bootstrapped MPLE}, that was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that this bootstrapped MPLE is a multivariate \textit{M}-estimator \cite{Huber1981}, a special class of robust estimators, meaning that this algorithm consistently estimates the confidence intervals of the MPLE.

\noindent We verify the consistency of the bootstrapped MPLE by conducting a simulation study on the same two networks with the same paramtrization as in the previous chapter: The Faux Mesa High friendship network and the Faux Magnolia High friendship network. \\[0.3cm]
For the simulation study, we determine the MPLE for the model and treat these estimates as the networks 'true' parameter values. We then use these parameter values to simulate a sample of $1000$ networks from the distribution of $Y$. For each of the $1000$ networks, we calculate 95\% confidence intervals based on the MCMCMLE and the logistic regression and examine whether the 'true' parameter values lie in these intervals. In addition, we determine the bootstrapped MPLE confidence intervals by sampling $500$ networks for each of the originally sampled $1000$ networks, by using the respective MPLE as parameter values. For every newly sampled network, we again determine the MPLE and then take the $2.5$th and $97.5$th percentile of the $500$ MPLE estimates to obtain a 95\% bootstrap confidence intervals. Similar as for the MCMCMLE and the logistic regression, we verify whether the 'true' parameter value can be found in the bootstrapped confidence interval.

\noindent Figure \ref{coverage} visualizes the coverage percentages for each of the three methods for both networks. The dashed line is set at $0.95$ and represents the optimal value. It is in evidence that the bootstrapped MPLE performed equally well as the MCMCMLE, achieving results that obtain the true parameter values in approximately 95\% of the cases. Additionally, a difference in the results between the smaller Faux Mesa High network and larger Faux Magnolia High network is not identifiable. Similar to the results of van Duijn et al \cite{vanDuijnetal2009} our results for the logistic regression differ distinctively from the anticipated 95\%, visualizing that the MPLE underestimates the variance of its estimates.
Figure \ref{bias} illustrates the bias between the 'true' network coefficients $\theta$ and the MPLE estimates. \textcolor{blue}{(=> not degenerate?, maybe provide boxplots for scnetwork)}
\noindent This simualtion study shows that bootstrapped MPLE is able to overcome the main disadvantage of the MPLE by retaining the validity of confidence intervals. Furthermore, this method also adopts the advantage of the MPLE of being consistent. However, the main advantage of the bootstrapped MPLE is the rapidity of the estimation that provide a computational benefit, especially if the size of the examined network is extremely large.

@@

\item describe advantage over MPLE (better coverage) and MCMCMLE (faster)
\item can easily be parallelized/ using multiple cores -> increases speed
\item this method is also able to detect degeneracy
\item include table with boot MPLE results and MCMCMLE results
\end{itemize}

 \section*{Other chapters}

 \begin{enumerate}
 \item Fields that made fruitful use of the MPLE for networks, e.g. Boltzmann machine
 \item Examples where a logit approach was used instead of an ERGM, because the network was too big and it was too time consuming using the MCMCMLE. We can then try to apply the bootstrapped MPLE approach.
\end{enumerate}

@@
